Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-22 paper/paper.tex
Final submission with updated benchmarks.

\emph{Variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}. These graphs are eventually expected to replace reference sequences as reference genomes. The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index used in vg. A \emph{path index} is a generalization of text indexes for labeled graphs. Given a query string, they find the paths with a label matching the query. Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length. The design of a path index is hence a trade-off between maximum query length, the size of the index, and pruning complex regions of the graph.

@@

The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently. Rødland \cite{Roedland2013} proposed another similar generalization. Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space usage by representing unary paths in the graph as single nodes. A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.

@@

Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

Suffix trees, suffix arrays, and \FMindex{} can be generalized to index multiple texts. There are also generalizations for other combinatorial structures. The \emph{XBW transform} \cite{Ferragina2009b} is essentially an \FMindex{} for \emph{labeled trees}. The nodes of the tree are sorted by path labels from the node to the root of the tree. $\BWT$ stores the labels of the children of each node (leaf nodes require special treatment). If a node has $k$ children, we encode that as binary sequence $0^{k-1} 1$. We concatenate these sequences to form bitvector $B$, which is used for mapping lexicographic ranks of nodes to the corresponding BWT ranges. The labels of the children of node $i$ are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$.

@@

Another representation trades query performance for the ability to search for short patterns. We sort the key-value pairs by their keys and store them in array $A$. If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of the key $X_{i}$, with binary search. We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.

We can use de~Bruijn graphs and \kmer{k} indexes to simulate each other. For that purpose, we define the de~Bruijn graph of graph $G = (V, E)$ by using the collection $\mathcal{S}$ of the labels of the maximal paths in the graph. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We transform $\mathcal{S}$ into a \kcollection{k} by inserting characters $\#$ to the beginning of each non-left-infinite sequence when necessary, and characters $\$$ to the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the start node and $j$ is the distance to the nearest non-inserted $\#$ in $S$. Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.

\begin{definition}[de~Bruijn graph of a graph]\label{def:dbg}
Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

\begin{definition}[Path graph as an index]
Let $G' = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
\begin{itemize}
\item Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$, if there is a path $P'$ in $G'$ with $G'.\glabel(P') = X$. We use $G'.\find(X)$ to denote the set of nodes $V'_{X} \subseteq V'$ matching the pattern.
\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = \bigcup_{v \in V'_{X}} G'.\gvalue(v)$. We use $G'.\locate(X)$ as a shorthand for $G'.\locate(G'.\find(X))$.
\end{itemize}
\end{definition}

@@

\begin{definition}[Equivalent path graphs]
Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
\end{definition}

@@

\begin{definition}[Maximally pruned de~Bruijn graph]
Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it using Lemma~\ref{lemma:dbg-prune} any further.
\end{definition}

@@

We now show how a GCSA of a path graph works. As path graphs are not exactly the same as the prefix-sorted automata used with the original GCSA \cite{Siren2014}, the technical details will be different. We also extend GCSA with techniques based on suffix trees.
See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
Appendix~\ref{appendix:construction} gives a construction algorithm for the GCSA of a pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph, starting from a set of paths of length $k$. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

The range of outgoing edges $(u', v') \in E'$ for node $u' \in V'$ with lexicographic rank $i'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$. The edges are already sorted by keys $G'.\gkey(u')$. Because graph $G'$ is a path graph, we know that key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. The sorting orders are therefore compatible. For every $j \in [sp_{in}, ep_{in}]$ for a node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for a node $u' \in V'$ implies an edge $(u', v') \in E'$. If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.

We use \emph{backward searching} for query $G'.\find(X)$. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes. Now assume that $\abs{X} \ge 1$ and that $G'.\find(X[i+1, \abs{X}-1]) = [sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range $G'.\find(X[i, \abs{X}-1])$, which is the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges; the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i+1}], X[i])$.

@@

We sample the values $G'.\gvalue(v')$ for a node $v' \in V'$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$ for the only incoming edge $(u', v') \in E'$. We may also sample the values for some nodes on long unary paths for performance reasons. If the values $G'.\gvalue(v')$ have not been sampled, we can derive them from sampled values by following the incoming edges.

If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$ . If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.

@@

Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys. We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree} of graph $G'$.

@@

BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference. It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions. GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$). However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.

@@

GCSA2 has been developed as the path index library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. We chose fast succinct data structures over the smaller but slower compressed data structures for most index components. The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0 for the experiments. All files were stored on a distributed Lustre file system. We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the queries.

@@

We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}. To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial with \texttt{vg mod -p -l -e 4} and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}. We extracted paths of length $16$ from the forward strands of all chromosomes. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes. The results are summarized in Table~\ref{table:construction}.  We could build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache. The de~Bruijn graph had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes. The index grew more slowly with $k$, as we sampled the values more often in lower order graphs.

@@

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{64} GCSA and an FM\nobreakdash-index. Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
\begin{tabular}{c|cccc|cccc}
\hline
 & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
GCSA: 16 & 351584 & 347453 & 1829M &  872M & 17.9 \textmu{}s & 0.51 \textmu{}s & 0.93 \textmu{}s & 10.3 \textmu{}s \\
GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.6 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 74.6 \textmu{}s & 0.25 \textmu{}s & 0.32 \textmu{}s & 1.65 \textmu{}s \\
\hline
FM:   16 & 351584 & 301538 &    -- &  782M & 7.83 \textmu{}s &              -- &              -- & 3.10 \textmu{}s \\
FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.16 \textmu{}s \\
FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.89 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i])$, and assume that it is nonempty. Edge $(u', v') \in E'$ exists if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. Hence $u' \in M_{i}$ if and only if key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$. If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$. Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
\end{proof}

@@

Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}. Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$. As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ for all nodes $v \in G'.\gvalue(v')$.

>>>>

\emph{Variation graphs} (also called pan-genomes, graph genomes, and reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}. These graphs are eventually expected to replace reference sequences as reference genomes. The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index used in vg. A \emph{path index} is a generalization of text indexes for labeled graphs. Given a query string, it finds the paths with a label matching the query. Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length. The design of a path index is hence a trade-off between maximum query length, the size of the index, and pruning complex regions of the graph.

@@

The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently. Rødland \cite{Roedland2013} proposed another similar generalization. Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space usage by representing unary paths in the graph as single nodes. A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.

@@

Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix, and \emph{right-maximal} if it is right-infinite or a suffix.

@@

Suffix trees, suffix arrays, and the \FMindex{} can be generalized to index multiple texts. There are also generalizations for other combinatorial structures. The \emph{XBW transform} \cite{Ferragina2009b} is essentially an \FMindex{} for \emph{labeled trees}. The nodes of the tree are sorted by path labels from the node to the root of the tree. $\BWT$ stores the labels of the children of each node (leaf nodes require special treatment). If a node has $k$ children, we encode that as a binary sequence $0^{k-1} 1$. We concatenate these sequences to form bitvector $B$, which is used for mapping lexicographic ranks of nodes to the corresponding BWT ranges. The labels of the children of the $i$th node are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$.

@@

Another representation trades query performance for the ability to search for short patterns. We sort the key-value pairs by their keys and store them in array $A$. If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of key $X_{i}$, with binary search. We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.

We can use de~Bruijn graphs and \kmer{k} indexes to simulate each other. For that purpose, we define the de~Bruijn graph of graph $G = (V, E)$ by using the collection $\mathcal{S}$ of the labels of the maximal paths in the graph. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We transform $\mathcal{S}$ into a \kcollection{k} by inserting characters $\#$ to the beginning of each non-left-infinite sequence when necessary, and characters $\$$ to the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the start node and $j$ is the distance to the nearest non-inserted $\#$ in $S$. Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = t:j$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.

\begin{definition}[De~Bruijn graph of a graph]\label{def:dbg}
Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

\begin{definition}[Path graph as an index]
Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
\begin{itemize}
\item Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$, if there is a path $P'$ in $G'$ with $G'.\glabel(P') = X$. We use $G'.\find(X)$ to denote the set of nodes $V'_{X} \subseteq V'$ matching the pattern.
\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = \bigcup_{v' \in V'_{X}} G'.\gvalue(v')$. We use $G'.\locate(X)$ as a shorthand for $G'.\locate(G'.\find(X))$.
\end{itemize}
\end{definition}

@@

\begin{definition}[Equivalent path graphs]
Let $G'$ and $G''$ be two path graphs, and let $k > 0$ be a parameter value. We say that graphs $G'$ and $G''$ are \emph{\kequivalent{k}}, if we have $G'.\locate(X) = G''.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
\end{definition}

@@

\begin{definition}[Maximally pruned de~Bruijn graph]
Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
\end{definition}

@@

We now show how a GCSA of a path graph works. As path graphs are not exactly the same as the prefix-sorted automata used with the original GCSA \cite{Siren2014}, the technical details will be different.
See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
We also extend GCSA with techniques based on suffix trees.
Appendix~\ref{appendix:construction} gives a construction algorithm for the GCSA of a pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph, starting from a set of paths of length $k$. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to that node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of the predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

The range of outgoing edges $(u', v') \in E'$ from node $u' \in V'$ with lexicographic rank $i'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$. The edges are already sorted by keys $G'.\gkey(u')$. Because graph $G'$ is a path graph, we know that key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. The sorting orders are therefore compatible. For every $j \in [sp_{in}, ep_{in}]$ for a node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for a node $u' \in V'$ implies an edge $(u', v') \in E'$. If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.

We use \emph{backward searching} for query $G'.\find(X)$. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes. Now assume that $\abs{X} \ge 1$ and that $G'.\find(X[i+1, \abs{X}-1]) = [sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range $G'.\find(X[i, \abs{X}-1])$, which is the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to the range $[sp_{in}, ep_{in}]$ of incoming edges; the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}] = G'.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i+1}], X[i])$.

@@

We sample the values $G'.\gvalue(v')$ for a node $v' \in V'$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$ for the only incoming edge $(u', v') \in E'$. We may also sample the values for some nodes on long unary paths for performance reasons. If the set $G'.\gvalue(v')$ has not been sampled, we can derive it from sampled values by following the incoming edges.

If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$. If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.

@@

Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys. We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree} of graph $G'$.

@@

BWA-MEM \cite{Li2013} aligns reads to a reference genome by finding \emph{maximal exact matches} between the read and the reference. It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions. GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$). However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.

@@

GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}. The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. The index uses fast succinct data structures instead of the smaller but slower compressed data structures in most of its components. The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0. All files were stored on a distributed Lustre file system. We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the queries.

@@

We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}. To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}. We extracted all paths of length $16$ from the forward strands of the graphs. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes. The results are summarized in Table~\ref{table:construction}.  We could build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache. The de~Bruijn graphs had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes. The index grew more slowly with $k$, as we sampled the values more often in lower order graphs.

@@

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{64} GCSA and an FM\nobreakdash-index. Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
\begin{tabular}{c|cccc|cccc}
\hline
 & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
GCSA: 16 & 351584 & 347453 & 1829M &  872M & 18.0 \textmu{}s & 0.40 \textmu{}s & 0.93 \textmu{}s & 8.61 \textmu{}s \\
GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.7 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 75.3 \textmu{}s & 0.25 \textmu{}s & 0.33 \textmu{}s & 1.66 \textmu{}s \\
\hline
FM:   16 & 351584 & 301538 &    -- &  782M & 6.04 \textmu{}s &              -- &              -- & 2.44 \textmu{}s \\
FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.17 \textmu{}s \\
FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.92 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i])$, and assume that it is nonempty. Edge $(u', v') \in E'$ exists if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. Hence if $u' \in M_{i}$, key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$. If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$. Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
\end{proof}

@@

Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}. Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$. As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.
